---
title: "Statistical Analysis of Flight Assay Data"
author: "Jun Ishigohoka"
output: 
  pdf_document:
    highlight: tango
fontsize: 12pt
header-includes:
 \usepackage{titling}
linkcolor: blue
urlcolor: blue

---




```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning=F, message = F, class.source = "numberLines lineAnchors")
knitr::opts_knit$set(root.dir = paste0(Sys.getenv("HOME"), "/Dropbox/work/conferences/2025/EMBO_Drosophila/prep/github/"))

```

\pagenumbering{gobble}

```{r, echo = F, fig.align = "center"}
img <- png::readPNG("pictures/cover.png")
grid::grid.raster(img)
```

\newpage
\pagenumbering{arabic}


# Preparation

If you don't have the packages, install them.

```{r}
packages <- c("ggplot2", "car")
for (pkg in packages) {
  if (!requireNamespace(pkg, quietly = TRUE)) {
    install.packages(pkg)
  }
}

```


Load libraries.

```{r}
library(ggplot2)
library(car) # Anova() and leveneTest()
```

Read data in two data frames.
For this protocol, test data will be used.


```{r}
d_1 <- read.csv("data/test_data/R4_female.csv")
d_2 <- read.csv("data/test_data/R4_male.csv")

head(d_1)
```

Add a column for the factor you are comparing.

```{r}
d_1$sex <- "female"
d_2$sex <- "male"
```

Concatenate the two data frames. Then set it as a factor.

```{r}
d <- rbind(d_1, d_2)
d$sex <- factor(d$sex, levels = c("female", "male"))

```

# Visualisation

Let's visualise the distribution of landing height.
**Once you have visualised the distributions, try to describe them.**

If you have `ggplot2`...

```{r}
ggplot(data = d, # data to be plotted
       mapping = aes(x=sex, y=Y, fill = sex)) + # columns in the data frame to be used for plotting
        geom_violin() +   # violin plot
        geom_boxplot(     # box plot
          width = 0.1,    # width of the box plot
          fill = "white", # colour inside the box plot
          outliers = F    # whether to have points for outliers. 
                          # FALSE because of geom_fitter() below
          ) +
        geom_jitter(      # each data point with "jitter" along x axis
          size = 0.5,     # size of points
          width = 0.1     # the amount of jitter
          ) + 
        scale_y_reverse(  # to make the plot same direction as the tapes
          limits = c(NA, 0) # so that the lowest Y is 0
          ) +
  labs(                   # Add labels
    x = "Sex", 
    y = "Landing position [mm]") +
  theme_bw()

```

If you prefer the base plot or do not have `ggplot2` installed.

```{r}
boxplot(d$Y ~ as.factor(d$sex),
     pch = NA,
     xlab = "Sex",
     ylab = "Landing position [mm]",
     ylim = rev(range(d$Y))
)
points(jitter(as.numeric(as.factor(d$sex))), d$Y)

```







# Statistical analysis (hypothesis testing)

## Are your data normally distributed?

Many parametric tests assume normality.
Let's run Shapiro-Wilk test for each group to test normality.

```{r}
shapiro.test(subset(d, sex == "female")$Y)
shapiro.test(subset(d, sex == "male")$Y)
```

Does the result fit your description of the distributions?

If they are not normally distributed, think why.
Do you expect normally distributed data from this experiment?
(Hints: Poisson process; waiting time)



## Does your data have equal variance?

Many parametric tests assume equal variance between groups.
Brown-Forsythe test for homogeneity. If you have `car` installed

```{r}
leveneTest(Y ~ sex, d, center = median)

```

If you do not have `car` installed, compute deviance from median for each sex manually, and run ANOVA.

```{r}
d$z <- NA
for (sex in levels(d$sex)) {
  idx <- which(d$sex == sex)
  d[idx,]$z <- abs(d[idx, ]$Y - median(d[idx, ]$Y))
}

summary(aov(d$z ~ d$sex))
```



## Is the mean landing distance different between groups?

If the data are normally distributed and have equal variance, Student t-test

```{r}
t.test(d$Y ~ d$sex,
       var.equal = T,
       paired = F
       )
```

If data are normal but not with equal variance, Welch's t-test

```{r}

t.test(d$Y ~ d$sex,
       var.equal = F,
       paired = F
       )


```

If data are not normally distributed, you can run non-parametric tests or permutation test.


**Option 1:** Non-parametric test

Let's run Mann–Whitney U test.

```{r}
wilcox.test(d$Y ~ d$sex,
            paired = F
            )
```



**Option 2**: Permutation test

Let's write a function to run a permutation test.
Briefly, we shuffle the label, compute difference between groups.
We repeat this process many times to get a null distribution.
We compute p-value based on the position of the observed difference between groups in the null distribution.

```{r}
# Null distribution of mean(Y | female) - mean(Y | male)
null_dist <- sapply(1:1000, # repeat 1000 times
       function(x){
         d_tmp <- d
         d_tmp$sex <- sample(d$sex) # shuffle label
         return(mean(d_tmp[d_tmp$sex == "female","Y"]) 
                  - mean(d_tmp[d_tmp$sex == "male","Y"])
                )
       }
       )

# Observed value of mean(female) - mean(male)
obs <- mean(d[d$sex == "female", "Y"] - d[d$sex == "male", "Y"])

# p value as the rank of observation in the null distribution
p_val <- sum(abs(null_dist) > abs(obs)) / length(null_dist)

hist(null_dist,
     main = paste("p_val: ", p_val))
abline(v = obs, col = "red", lty = 2)
```

# Optional: Statistical modeling

Here, we conduct statistical modeling of landing distance using a generalised linear model (GLM) approach with the exponential distribution, considering landing events as a Poisson process with a constant "landing rate".
Because the waiting time of a Poisson process is an exponentially distributed random variable, we can use a GLM with exponential distribution.
We compare the result with a linear model (LM), which assumes a normal distribution as the underlying distribution.

## GLM


Fit the data to an exponential GLM.

```{r}
glm_1 <- glm(Y ~ sex,
             data = d, 
             family = Gamma(link = "log") # This is how to specify exponential
             )
```

**According to the fitted GLM, do the two groups have different landing rate?**

If you have `car` installed

```{r}
Anova(glm_1)
```

```{r}
summary(glm_1, dispersion = 1) # exponential is a Gamma with dispersion = 1
```

If you do not have `car` installed

```{r}
drop1(glm_1, test = "Chi")
```

## LM

Fit the data to a linear model.

```{r}
lm_1 <- lm(Y ~ sex,
             data = d)
```

If you do not have `car` installed

```{r}
drop1(lm_1, test = "Chi")
```

```{r}
summary(lm_1)
```

## GLM vs LM

Compare the diagnostic plots[^1] between `glm_1` and `lm_1`.
**Is using LM for landing distance data worse than exponential GLM?"

[^1]: What do the diagnostic plots mean? <https://www.sthda.com/english/articles/39-regression-model-diagnostics/161-linear-regression-assumptions-and-diagnostics-in-r-essentials/>

-   Is the linearity assumption met?
-   Is the data homoscedastic?
-   Are residuals distributed normally?
-   Are there outliers?

```{r, fig.height = 7, fig.width = 7}
par(mfrow = c(2,2),
    oma = c(0, 0, 2, 0)
    )
plot(lm_1)
```

```{r, fig.height = 7, fig.width = 7}
par(mfrow = c(2,2),
    oma = c(0, 0, 2, 0)
    )
plot(glm_1)
```

## What's next?

In real life, we have multiple biological and technical replicates.
**How do we account for such random effects?**




```{r}
sessionInfo()
```


